---
title: "Time"
keypoints:
- Use the `dplyr` package to manipulate dataframes.
- Use `select()` to choose variables from a dataframe.
- Use `filter()` to choose data based on values.
- Use `group_by()` and `summarize()` to work with subsets of data.
- Use `mutate()` to create new variables.
- Use the `tidyr` package to change the layout of dataframes.
- Use `pivot_wider()` to go from long to wide format.
- Use `pivot_longer()` to go from wide to long format.
objectives:
- Describe the purpose of an R package and the **`dplyr`** and **`tidyr`** packages.
- Select certain columns in a dataframe with the **`dplyr`** function `select`.
- Select certain rows in a dataframe according to filtering conditions with the **`dplyr`**
  function `filter`.
- Link the output of one **`dplyr`** function to the input of another function with
  the 'pipe' operator `%>%`.
- Add new columns to a dataframe that are functions of existing columns with `mutate`.
- Use the split-apply-combine concept for data analysis.
- Use `summarize`, `group_by`, and `count` to split a dataframe into groups of observations,
  apply a summary statistics for each group, and then combine the results.
- Describe the concept of a wide and a long table format and for which purpose those
  formats are useful.
- Describe the roles of variable names and their associated values when a table is
  reshaped.
- Reshape a dataframe from long to wide format and back with the `pivot_wider` and
  `pivot_longer` commands from the **`tidyr`** package.
- Export a dataframe to a csv file.
questions:
- How can I select specific rows and/or columns from a dataframe?
- How can I combine multiple commands into a single command?
- How can I create new columns or remove existing columns from a dataframe?
- How can I reformat a dataframe to meet my needs?
teaching: 50
exercises: 30
source: Rmd
---


```{r, include = FALSE}
source("../bin/chunk-options.R")
knitr_fig_path("03-")
source("../bin/download_data.R")
```


A relatively short session on time.


“People assume that time is a strict progression from cause to effect, but actually from a non-linear, non-subjective viewpoint, it's more like a big ball of wibbly-wobbly, timey-wimey stuff.”

Time is not easy to deal with. It is actually really complicated. Here is a rant on how complicated it is...
https://www.youtube.com/watch?v=-5wpm-gesOY

Here, we are not going to spend a lot of time on it. Statistics Denmark returns a
lot of data-series by quarter, or month. And we need to convert it to something 
we can work with. Withou necessarily understanding all the details.

Two packages will offer a lot of help.
 
lubridate and tsibble

They may need to be installed first.
After that, we can load them:
```{r}
library(lubridate)
library(tsibble)
```

Lubridate converts a lot of different ways of writing dates to a consistent 
date-time format.

The most important functions we need to know, are:

ymd
hms
ymd_hms

And variations of these, especially ymd.

ymd("2021-09-21") converts the date 2020-09-21 to a date-format that R can understand.

Sometimes we have dates formatted as "21-09-2021". Or "09-21-2021" or
"21-09-21"




```{r, results = 'hide', purl = FALSE, message = FALSE}

## load the tidyverse
library(tidyverse)
library(here)

interviews <- read_csv(here("data", "SAFI_clean.csv"), na = "NULL")

## inspect the data
interviews

## preview the data
# view(interviews)
```



```{r, purl = FALSE}
interviews2 <- filter(interviews, village == "Chirodzo")
interviews_ch <- select(interviews2, village:respondent_wall_type)
```

```{r, purl = FALSE}
interviews_ch <- select(filter(interviews, village == "Chirodzo"),
                         village:respondent_wall_type)
```



```{r, purl = FALSE}
interviews_ch <- interviews %>%
    filter(village == "Chirodzo") %>%
    select(village:respondent_wall_type)

interviews_ch

```




{% include links.md %}
