---
title: "What is an API?"
teaching: 30
exercises: 15
output: 
  html_document:
    df_print: paged
objectives:
- Create a .Rmd document containing R code, text, and plots
- Create a YAML header to control output
- Understand basic syntax of (R)Markdown
- Customise code chunks to control formatting
- Use code chunks and in-line code to create dynamic, reproducible documents
keypoints: 
- R Markdown is a useful language for creating reproducible documents combining text and executable R-code.
- Specify chunk options to control formatting of the output document
source: Rmd
questions:
- What is R Markdown?
- How can I integrate my R code with text and plots?
- How can I convert .Rmd files to .html?
---

```{r setup, include = FALSE}
source("../bin/chunk-options.R")
knitr_fig_path("06-")
source("../bin/download_data.R")
```

## What is an API?

An API is an Application Programming Interface. It is a way of making 
applications, in our case an R-script, able to communicate with another 
application, here the Statistics Denmark databases.

Talking about APIs, we talk about several different things. It can be quite 
confusing, but dont worry!

What we want to be able to do, is to let our own application, our R-script,
send a command to a remote application, the databases of Statistics Denmark, 
in order to retrieve specific data.

An API defines the different commands we can send, and how the data that we 
get back, is formatted.

Often APIs will require a user account with a login and a password. 
Statistics Denmark does not.

The standard way to send a command, or a request, to an API is to use the 
GET (and POST) functions at the core of the internet. 

In a certain sense this is what we do when we access a website. We go to 
www.dr.dk/sporten and get a result, the current webpage at the front of 
the sports section of Danmarks Radio. 

If we instead ask www.dr.dk to return the result of our request for 
www.dr.dk/nyheder/politik, we will get the current webpage with news on politics.

This is what we do when we access an API. But instead of using our browser, we
use the method our browser uses (GET), tells that method that we would like 
some specified information, and get a result that is not a webpage, but rather
a set of data. Hopefully organised in a way that is easy to read.

Writing our own GET-requests to communicate with an API is not simple. 
Thankfully kind people have written libraries, some in R, that makes accessing 
specific APIs easier. The one we are going to use here is called "danstat"

## The danstat package/library


Before doing anything else, it is useful to take a look at the result:
```{r echo = F}
library(danstat)
variables <- list(list(code = "ieland", values = c(5100, 5128)),
                  list(code = "køn", values = c(1,2)),
                  list(code = "tid", values = NA))

data <- get_data("folk1c", variables)
data %>% head
```

This is from the table "folk1c" from Statistics Denmark.

We get some variables, IELAND, KØN, and TID. And then the content of the table, 
INDHOLD. Ie the number of men, living in denmark i the first quarter of 2008 in
the first line.

How do we get that table?

All tables from Statistics Denmark are organised in a hierarcical tree of 
subjects. 

Let us begin there.

Before using the library, we need to install it:

```{r eval=F}
install.packages("danstat")
```

After installation, we load the library using the library function. And then 
we can access the functions included in the library:

```{r}
library(danstat)
subjects <- get_subjects()
subjects
```

We get the 13 major subjects from Statistics Denmark. Each of them have sub-subjects.

Let us try to get the sub-subjects from the subject 02 - containing information
about populations and elections:

```{r}
sub_subjects <- get_subjects(subjects = "02")
sub_subjects
```



The get_subjects() function sends a request to the Statistics Denmark API, asking
for a list of the subjects. The information is returned to our script, and the
get_subjects() function presents us with a dataframe containing the information.

Each subject (may) have sub-subjects. If we want to take a closer look at the 
subdivisions of a given subject, we use the get_subjects() function again,
this time specifying which subject we are interested in:

```{r}
subject <- get_subjects("02")
subject
```


The result is a bit complicated. The column "subjects" in the resulting dataframe
contains another dataframe. We access it like we normally would access a 
column in a dataframe:

```{r}
subject$subjects
```

Those sub-subjects have their own subjects! Lets get to the bottom of this, and
use 2401, Population and population projections as an example:

```{r}
sub_sub_subjects <- get_subjects("2401")
sub_sub_subjects$subjects
```
Now we are at the bottom. We can see in the column "hasSubjects" that there 
are no sub_sub_sub_subjects. 

The hierarchy is:
02 Population and elections
| 
2401	Population and population projections
|
10021	Population in Denmark

The final sub_sub_subject contains a number of tables, that actually 
contains the data we are looking for.

get_subjects is able to retrieve all the sub, sub-sub and sub-sub-sub-jects in
one go. The result is a bit confusing and difficult to navigate.

Remember that the initial result was a dataframe containing another dataframe.
If we go all the way to the bottom, we will get a dataframe, containing several
dataframes, each of those containing several dataframes. 

We recommend that you do not try it, but this is how it is done:
```{r}
lots_of_subjects <- get_subjects("02", recursive = T, include_tables = T)
```

The "recursive = T" parameter means that get_subjects will retrieve 
the subjects of the subjects, and then the subjects of those subjects.

## Which datatables exists?

But we ended up with a sub_sub_subject, 

10021	Population in Denmark

How do we find out which tables exists in this subject?

The get_tables() function returns a dataframe with information about the 
tables available for a given subject.


```{r}
tables <- get_tables(subjects="10021")
tables
```

We get at lot of information here. The id identifies the table, text gives a 
description of the table that humans can understand. When the table was last
updated and the first and last period that the table contains data for.

In the variables column, we get information on what kind of data is stored in 
the table.

Before we pull out the data, we need to know which variables are available
in the table. We do this with this function:

```{r}
metadata <- get_table_metadata("FOLK1A", variables_only = T)
metadata
```

There is a lot of other metadata in the tables, including the phone number to 
the staffmember at Statistics Denmark that is responsible for maintaining the
table. We are only interested in the variables, which is why we add the 
parameter "variables_only = T".


What kind of values can the individual datapoints take?

```{r}
metadata %>% slice(4) %>% pull(values)
```

We use the slice function from tidyverse to pull out the fourth row of the 
dataframe, and the pull-function to pull out the values in the values
column.

The same trick can be done for the other fields in the table:

```{r}
metadata %>% slice(1) %>% pull(values) %>% .[[1]] %>% head
```
Here we see the individual municipalities in Denmark. 

Now we are almost ready to pull out the actual data!

But first!

## Which variables do we want?

We need to specify which variables we want in our answer. Do we want 
the total population for all municipalities in Denmark? Or just a few?
Do we want the total population, or do we want it broken down by sex.

These variables, and the values of them, need to be specified when we 
pull the data from Statistics Denmark.

We also need to provide that information in a specific way.

If we want data for all municipalites, we want to pull the variable 
"OMRÅDE" from the list of variables.

Therefore we need to give the function an argument containing both
the information that we want the population data broken down by "OMRÅDE", and 
that we want all values of "OMRÅDE".


Vectors are characterized by only being able to contain one type of data.

When we need to have structures that can contain more than one type of data,
we can use the list structure.

Lists allows us to have values, with names (sometime descriptive).

Lists can even contain lists.

And that is what we need here. Let us make our first list:

```{r}
list(code = "OMRÅDE", values = NA)
```
  
This list have to components. One called "code", and one called "values".
Code have the content "OMRÅDE", specifying that we want the variable in the 
data from Statistics Denmark calld "OMRÅDE".

"values" has the content "NA". We use "NA", when we want to specify that we
want all the "OMRÅDE". If we only wanted a specific municipality, we could 
instead specify it instead of writing "NA".

Let us assume that we also want to break down the data based on marriage status.

That information is stored in the variable "CIVILSTAND".

And above, we saw that we had the following values in that variable:
```{r}
metadata %>% slice(4) %>% pull(values)
```

A value for the total population is probably not that interesting, if we 
pull all the individual values for "Never married" etc.

We can now make another list:
```{r}
  list(code = "CIVILSTAND", values = c("U", "G", "E", "F"))
```

Here the "values" part is a vector containing the values we want to pull out 
for that variable.

It might be interesting to take a look at how the population changes over time.

In that case we need to pull out data from the "Tid" variable.

That would look like this:
```{r}
list(code = "Tid", values = NA)
```


If we want to pull data broken down by all three variables, we need to provide a
list, containing three lists. 

We do that using this code:


```{r}
variables <- list(list(code = "OMRÅDE", values = NA),
                  list(code = "CIVILSTAND", values = c("U", "G", "E", "F")),
                  list(code = "Tid", values = NA)
              )
variables
```

And now, finally, we are ready to get the data!

```{r}
data <- get_data(table_id = "FOLK1A", variables = variables)
```
It takes a short moment. But now we have a dataframe containing the data we 
requested:

```{r}
head(data)
```

This procedure will work for all the tables from Statistics Denmark!

The data is nicely formatted and ready to use. Almost.

Before we do anything else, let us save the data.
```{r}
write_csv2(data, "../data/SD_data.csv")
```

