---
title: "Time"
keypoints:

- Use `pivot_longer()` to go from wide to long format.
objectives:
- Describe the purpose of an R package and the **`dplyr`** and **`tidyr`** packages.
- Select certain columns in a dataframe with the **`dplyr`** function `select`.
questions:
- How can I select specific rows and/or columns from a dataframe?
teaching: 50
exercises: 30
source: Rmd
---


```{r, include = FALSE}
source("../bin/chunk-options.R")
knitr_fig_path("03-")
source("../bin/download_data.R")
```


# A relatively short session on time.


“People assume that time is a strict progression from cause to effect, but actually from a non-linear, non-subjective viewpoint, it's more like a big ball of wibbly-wobbly, timey-wimey stuff.”

Time is not easy to deal with. It is actually really complicated. Here is a rant 
on how complicated it is...

https://www.youtube.com/watch?v=-5wpm-gesOY


## Why?

We just pulled data out giving us the danish population, broken down by
marriage status and geographical area. And time.

If the data is not still in memory, we can read it in:

```{r}
data <- read_csv2("../data/SD_data.csv")
```


```{r}
head(data)
```

Note that the datatype for "TID" is <chr>, meaning character. Those are 
simply text, not a time. And if we want to plot this, as a function of time,
the "TID" variable needs to be converted into something R can understand as time.

## A general tool
lubridate is a package written to make working with dates and times easy(er).

It may need to be installed first.

```{r eval=F}
install.packages("lubridate")
```


After that, we can load it:
```{r}
library(lubridate)
```

Lubridate converts a lot of different ways of writing dates to a consistent 
date-time format.

The most important functions we need to know, are:

* ymd
* hms
* ymd_hms

And variations of these, especially ymd.

ymd("2021-09-21") converts the date 2020-09-21 to a date-format that R can 
understand:

```{r}
ymd("2021-09-21")
```


Sometimes we have dates formatted as "21-09-2021". That is day, month and year
in that order.

That can be converted to at standard date-format with the function dmy():
```{r}
dmy("21-09-2021")
```
We might even have dates formatted as "2021 21 4", (year, day month), the
function ydm() can handle that.
```{r}
ydm("2021 21 4")
```

Time is handled in a similar way, but time is usually not written as creatively
as dates:


```{r}
hm("14:05")
```
```{r}
hms("14.05.21")
```

Dates and times can be combined, as in: "2021-04-21 14:05:12":
```{r}
ymd_hms("2021-04-21 14:05:12")
```
Those were the nice dates...

## Not so nice date formats - a more specific tool

Statistics Denmark returns a lot of data-series by quarter, or month. 
And we need to convert it to something 
we can work with. Without necessarily understanding all the details.

The library tsibble provides functions that can convert "2020Q1", the first 
quarter of 2020, into something R can understand as time-value:

We might need to install it first:
```{r eval=F}
install.packages("tsibble")
```

And then load it:
```{r}
library(tsibble)
```

This is a vector containg the 8 quarters of the years 2019 and 2020.

```{r}
quarters <- c("2019Q1", "2019Q2", "2019Q3", "2019Q4", "2020Q1", "2020Q2", "2020Q3", "2020Q4")
class(quarters)
```
It is a character vector, ie strings. If we want to analyse any data 
associated with these specific quarters, we need to convert them to something 
R is able to recognize as time.

```{r}
yearquarter(quarters)
```
We are not going to go into further details on the challenges of working with
time-series. The generic lubridate functions and yearquarter() will be enough for
our purposes.

Let us finish by converting the "TID" column in our data, to a time-format.
```{r}
data <- data %>% 
  mutate(TID = yearquarter(TID))
```

We mutate the column "TID" into the result of running yearquarter() on the
column "TID". And now we have a data frame that we can do interesting things 
with.

Now might be a good time to save the data in its new version:
```{r eval=F}
write_csv2(data, "../data/SD_data.csv")
```

Note that we are using write_csv2() here. We do not have decimalpoints in this
data, but other data might have.

{% include links.md %}
